The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-01-12 paper/ms.tex
Almost

@@

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions the average number of
scatterings has a modest change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In the same set-up the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the homogeneous
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In the spheres with centrally
distributed sources we find $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depths
$\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$

@@

The top right panel of Figure \ref{fig:Nscatt2D} supports our hypothesis about
the photo-sphere in the homogeneous distribution. In that case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This explains the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, in the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits where $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. First of all we note that the escape fraction does not
increase from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, this is
explained since there is a transition from a opaque to an extremely opaque medium
which affect affects how \ly photons scape from the medium: single flight 
vs single excursion.
 %we are in a regime where the condition for the analytic expectations  ($a\tau_{h}\gg 1$) does not hold. 
This figure also shows that the escape fraction is higher by factors of $2-10$
than the expectation for a slab with centrally distributed
sources. Furthermore, the decrease of the escape fraction with
increasing optical depth is also slower.

transition from the opaque to
  extremely opaque medium, which affects how Lya photons escape from
  the medium: single flight vs single excursion

@@

Figure \ref{fig:OCspheres} shows the two spherical regions we
define to test for this effect. Each has a radius of $0.5$R and can
have two possible centers positions $C$ along the $\hat{\j}$ direction:
$\pm C\hat{\i}$. We also place an observer along the $\hat{\j}$
direction, such that the emitting sphere centered on $+C\hat{\j}$ moves
away from the observer and the sphere centered on $+C\hat{\j}$ moves
towards the observer.

@@

The variations in the $\mu$ distribution is shown in Fig.\ref{fig:muhisto},
for the Off-Center model there is a strong dependence of the 
observed flux on the relative position of an observer with $\Delta\mu\leq 25\%$.
In contrast,the central and homogeneous source distributions only show a 
weak anisotropy at the level of $\Delta\mu\leq 3\%$. The variations
in $\mu$ does not depend on the rotational velocity while increasing the optical depth $\tau_{H}$
would also increase the variation of $\Delta\mu$ up to a $15\%$.

Figure \ref{fig:OC_profiles} show the spectra for the two off-centered
spheres located at $C=\pm0.5R$.  In this case we only use photons that
have outgoing directions aligned with the observer i.e $\vert \mu \vert > 
0.9$ $\&$ $\vert \mu \vert < 1.0]$.  We do not find any asymmetry in 
the line produced by the off-center emission, the difference
 between the profiles corresponding to the spheres located at $C=-0.5R$ 
 and that ones located at $C=0.5R$ is negligible.  As in the homogeneous / 
central model the line gets broader as the rotational velocity increase also 
the central part of the line becomes predominant . 
There is also a signature of possible three peaks profiles for some models ($\tau = 10^6,
V=100,200$\kms and $\tau = 10^7, V=200,300$\kms) which would be studied 
in detail in the next section.

@@

As mentioned above Off-centered emission shows signs of three peak 
profiles, to analyse this in more detail we use the same method of the spheres
explained in \S \ref{sec:off-center}. We select spheres at $C=(\pm 0.25R, \pm0.5R, \pm0.75R)$ with a radii of
$0.25R$ for all the models, for those who present signs of three peak profiles such as 
$(V_{max}=50km$ $s^{-1}$ \& $\tau_{H}=10^{5}), (V_{max}=100km$ $s^{-1}$ \& $\tau_{H}=10^{6}), 
(V_{max}=200km$ $s^{-1}$ \& $\tau_{H}=10^{7})$ we made another run with $\tau_{\gamma}=10^{6}$ 
with the aim of get a stronger profile.

Fig. \ref{fig:3p_profiles} shows our findings, in order to get such profiles 
a relation between optical depth, rotational velocity and $C$ must be accomplish.
In particular we found that for the outer spheres $C=\pm 0.75$ the three
peak arises for all of our models with $\tau_{\gamma}=10^{6}$ this is due to 
the fact that the wings of the line move away from the central part as $C$ increase. 
Another requirement is that a high optical depth must be balanced with a high velocity 
in order to make the wings and the central peak similar in 
height, for models in which the velocity is low spheres at $C = \pm 0.25$ produce 
noise three peak profiles.

Fig.\ref{fig:3p_res} show that even with a low resolution a three-peak profile 
would be observed, this is important for observational porpoises due to the finite 
spectral resolution.

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion focuses on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} ans \cite{Yamada2012}

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation.

@@

\begin{itemize}
\item For central sources we find that the number of scatterings
does not significantly decrease with rotation, this is due to the 
fact that the vast majority of scatterings events are resonant 
during which the mean free path is very short and the effect of gas
bulk rotation is not enough to affect \ly photons.
The escape fraction does not depend on the rotational velocity neither
the position of the peaks, but there is an increase 
in the flux on the central parts of the line for high rotational 
velocities.

\item For homogeneous sources we find the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings and on average 
with $40\%$ less scatterings than in the static case. This makes highly improbable 
that a photon's frequency can move away from the line's center. 
In \S \ref{sec:widthpeak} For typical column densities of $N_{HI}=1e^{19}-1e^{20}$, 
we find that rotational velocities of 100-300 \kms broaden the line profile up 
to a factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed. Also the escape fraction increase by factors of $2-10$ with 
respect to the slab with centrally sources.

\item For non-symmetric sources we find that three peak profiles
could emerge for models in which the rotational velocity is the necessary 
to give rise to a central peak but also not that high to decrease the
amount of photons in the wings. When initial off-center sources are 
selected from the outer parts of the galaxy the wings of the line are stronger.
Also off-center emission do not produce any considerable asymmetries in the line.
\end{itemize}

@@

JNGC thanks the support of Universidad de los Andes given as teaching assistant, 
Also to the  ISAAC 2012 summer school in San Diego which provide computer time
 where some of this models where ran.

>>>>

% No entiendo bien esta ultima frase, la comparasion es entre que modelos? con que \tau_H?. Nicolas

@@

% No entiendo la restricción!

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions the average number of
scatterings has a modest change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In the same set-up the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the central
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In the spheres with centrally
distributed sources we find $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depths
$\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$

@@

The top right panel of Figure \ref{fig:Nscatt2D} supports our hypothesis about
the photo-sphere in the homogeneous distribution. In that case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This explains the origin of a single central
peak. However, for a central distribution the situation is
different (lower panels). In this case the number of scatterings remains high, in the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits where $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. First of all we note that the escape fraction does not
increase from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, this is
explained since there is a transition from a opaque to an extremely opaque medium
which affect affects how \ly photons escape from the medium: single flight 
vs single excursion.
 %we are in a regime where the condition for the analytic expectations  ($a\tau_{h}\gg 1$) does not hold. 
This figure also shows that the escape fraction is higher by factors of $2-10$
than the expectation for a slab with centrally distributed
sources. Furthermore, the decrease of the escape fraction with
increasing optical depth is also slower.

@@

Figure \ref{fig:OCspheres} shows the two spherical regions we
define to test for this effect. Each has a radius of $0.5$R and can
have two possible centers positions $C$ along the $\hat{\j}$ direction:
$\pm C\hat{\i}$. We also place an observer along the $\hat{\j}$
direction, such that the emitting sphere centered on $+C\hat{\j}$ moves
away from the observer and the sphere centered on $-C\hat{\j}$ moves
towards the observer.

@@

The variations in the $\mu$ distribution is shown in Figure \ref{fig:muhisto},
for the Off-Center model there is a strong dependence of the 
observed flux on the relative position of an observer with $\Delta\mu\leq 25\%$.
In contrast,the central and homogeneous source distributions only show a 
weak anisotropy at the level of $\Delta\mu\leq 3\%$. The variations
in $\mu$ does not depend on the rotational velocity while increasing the optical depth $\tau_{H}$
would also increase the variation of $\Delta\mu$ up to a $15\%$.

Figure \ref{fig:OC_profiles} show the spectra for the two off-centered
spheres located at $C=\pm0.5R$.  In this case we only use photons that
have outgoing directions aligned with the observer i.e $\vert \mu \vert > 
0.9$ $\&$ $\vert \mu \vert < 1.0]$.  We do not find any asymmetry in 
the line produced by the off-center emission, the difference
 between the profiles corresponding to the spheres located at $C=-0.5R$ 
 and that ones located at $C=0.5R$ is negligible.  As in the homogeneous/
central model the line gets broader as the rotational velocity increase also 
the central part of the line becomes predominant . 
There is also a signature of possible three peaks profiles for some models ($\tau = 10^6,
V=100,200$\kms and $\tau = 10^7, V=200,300$\kms) which would be studied 
in detail in the next section.

@@

As mentioned above Off-centered emission shows signs of three peak 
profiles, to analyse this in more detail we use the same method of the spheres
explained in \S \ref{sec:off-center}. We select spheres at $C=(\pm 0.25R, \pm0.5R, \pm0.75R)$ with a radii of
$0.25R$ for all the models, for those who present signs of three peak profiles such as 
$(V_{max}=50km$ $s^{-1}$ \& $\tau_{H}=10^{5}), (V_{max}=100km$ $s^{-1}$ \& $\tau_{H}=10^{6}), 
(V_{max}=200km$ $s^{-1}$ \& $\tau_{H}=10^{7})$ we made another run with $\tau_{\gamma}=10^{6}$ 
with the aim of get stronger profiles.

Figure \ref{fig:3p_profiles} shows our findings, in order to get such profiles 
a relation between optical depth, rotational velocity and $C$ must be accomplish.
In particular we found that for the outer spheres $C=\pm 0.75$ the three
peak arises for all of our models with $\tau_{\gamma}=10^{6}$ this is due to 
the fact that the wings of the line move away from the central part as $C$ increase. 
Another requirement is that a high optical depth must be balanced with a high velocity 
in order to make the wings and the central peak similar in 
height, for models in which the velocity is low spheres at $C = \pm 0.25$ produce 
noise three peak profiles due to the high value of $\tau_{H, r}$.

Figure \ref{fig:3p_res} show that even with a low resolution a three-peak profile 
would be observed, this is important for observational porpoises due to the finite 
spectral resolution.

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion focuses on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} and \cite{Yamada2012}

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect that can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

@@

\begin{itemize}
\item For central sources we find that the number of scatterings
does not significantly decrease with rotation, this is due to the 
fact that the vast majority of scatterings events are resonant, 
during which the mean free path is very short and the effect of gas
bulk rotation is not enough to affect \ly photons.
The escape fraction does not depend on the rotational velocity neither
the position of the peaks, but there is an increase 
in the flux on the central parts of the line for high rotational 
velocities.

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings and on average 
with $40\%$ less scatterings than in the static case. This makes highly improbable 
that a photon's frequency can move away from the line's center. 
In \S \ref{sec:widthpeak} For typical column densities of $N_{HI}=1e^{19}-1e^{20}$, 
we find that rotational velocities of 100-300 \kms broaden the line profile up 
to a factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed. Also the escape fraction increase by factors of $2-10$ with 
respect to the slab with centrally sources.

\item For non-symmetric sources we find that three peak profiles
could emerge for models in which the rotational velocity is the necessary 
to give rise to a central peak, but not that high to decrease the
amount of photons in the wings. When initial off-center sources are 
selected from the outer parts of the galaxy the wings of the line are stronger.
Also we do not find any considerable asymmetries in the line produced by off-center 
emission.
\end{itemize}

@@

JNGC thanks the support of Universidad de los Andes given as teaching assistant, 
Also to the  ISAAC 2012 summer school held University of California San Diego which provide computer time
 where some of this models where ran.
